### 2020-07-05 chapters/abstract.tex
Resumo tuga

%Approach/Solution and %Main contributions
In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data. We develop a two-phased and two-windowed method: first, we perform batch analysis on a reference window and then a stream analysis over a sliding streaming window. We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages. We do this for each event and each of its fields also denominated features in our context. We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used. For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features. We generate alerts based on a user-defined probability threshold, for each feature.

We evaluate our method through a series of tests, some with synthetic datasets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next. Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity regarding window size.

\vspace*{10mm}\noindent
\textbf{Keywords}: streaming, real-time, lightweight, monitoring, concept drift, anomaly detection

>>>>

%Approach/Solution and %Main contributions
In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data. We develop a two-phased and two-windowed method: first, we perform batch analysis on a reference window and then a stream analysis over a sliding streaming window. We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs). We do this for each event and each of its fields, also denominated features in our context. We measure the divergence between both reference and sliding window EMA-based histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used. For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features. We generate alerts based on a user-defined probability threshold, for each feature.

We evaluate our method through a series of tests, some with synthetic datasets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next. Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.

\vspace*{10mm}\noindent
\textbf{Keywords}: data streams, monitoring, real-time, lightweight, concept drift

### 2020-07-05 chapters/resumo.tex
Resumo tuga

\chapter*{Resumo}

\vspace*{10mm}\noindent
\textbf{Keywords}:

>>>>

\chapter*{Resumo}
Muitos sistemas de monitorização de \textit{streams} de dados em tempo real são estáticos uma vez colocados em produção. Os engenheiros desses sistemas configuram-nos sob o pressuposto de que dados futuros que fluem pelo sistema seguirão aproximadamente a mesma distribuição que os dados vistos anteriormente. Fazem-no conscientemente, com as ferramentas disponíveis, tendo em mente que no futuro poderão precisar de reconfigurar o sistema. Assim, mesmo que a configuração inicial do sistema seja uma das melhores, ao longo do tempo, devido a mudanças nos padrões de dados, o desempenho do sistema em produção decai gradualmente. A detecção de mudança de padrões de dados refere-se ao processo de encontrar padrões em dados que não estão em conformidade com o comportamento esperado ou usual. Uma detecção oportuna e precisa dos desvios dos padrões de dados permite que medidas imediatas sejam tomadas. Assim, o problema em questão é determinar quando reconfigurar o sistema, \textit{e.g.}, um modelo de \textit{Machine Learning}, com base numa análise dos desvios de padrões da \textit{stream} de dados.

%Approach/Solution and %Main contributions
Nesta tese, projetamos um método que utiliza agregações de \textit{streaming} leves e funções de divergência de distribuição para alertar sobre desvios nos padrões de dados em tempo real, na presença de grandes volumes de dados a alta velocidade, distorcidos e sazonais. Desenvolvemos um método de duas fases e duas janelas: primeiro, realizamos uma análise \textit{batch} numa janela de referência e, em seguida, uma análise \textit{streaming} sobre uma janela deslizante. Agregamos o conteúdo de ambas as janelas usando um histograma aproximado baseado em Médias Móveis Exponenciais (MMEs). Fazemos isto para cada evento e cada um dos seus campos, também denominados \textit{features} no nosso contexto. Medimos a divergência entre os histogramas de referência e os das janelas deslizantes baseados em MMEs para cada \textit{feature} com a função de divergência de \textit{Jensen-Shannon}, mas outras funções podem ser usadas. Para cada \textit{feature}, descobrimos o valor de probabilidade para cada medida de divergência e aplicamos a correção de múltiplo teste de Holm-Bonferroni para descobrir quais os valores de probabilidade mais baixos e as \textit{features} associados. Geramos alertas  para cada \textit{feature} com base num limite de probabilidade definido pelo administrador.

Avaliamos o nosso método através de uma série de testes, alguns com conjuntos de dados sintéticos e outros com dados reais. Dividimos ainda os testes em testes de análise a uma única \textit{feature} e tests de análise com múltiplas \textit{features}. O nosso método detectou com precisão as anomalias introduzidas nos testes com conjuntos de dados sintéticos, mantendo um alto \textit{throughput}, tanto para análises com uma única ou múltiplas \textit{features}. No entanto, os testes com dados reais não foram tão precisos. Apesar de não sabermos as razões específicas para isso, passamos os próximos testes para trabalhos futuros e formulamos um conjunto de hipóteses que pretendem explicar o sucedido e, portanto, são dignas de serem exploradas em seguida. O nosso conjunto de testes apoia a afirmação de que métodos de agregação de janelas deslizantes e métodos de divergência de distribuição podem ser combinados para detectar alterações nos padrões de dados em cenários de \textit{streaming} com complexidade temporal e de memória constante.

\vspace*{10mm}\noindent
\textbf{Keywords}: data streams, monitoring, real-time, lightweight, concept drift
